---
title: "Prelminary field trial analysis"
author: "wolfemd"
date: "2022-04-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

Follow this worked example _and_ reference the 
["Genomic Selection Manual" - Chapter 9](https://wolfemd.github.io/GenomicSelectionManual/preliminary-field-trial-analysis.html).

# Load packages
```{r}
suppressMessages(library(tidyverse)); 
suppressMessages(library(genomicMateSelectR)); 
suppressMessages(library(sommer));
```

# Set-up the training data

```{r}
phenos<-readRDS(here::here("output","phenotypes_cleaned.rds"))
```

```{r}
phenos %>% 
     count(CompleteBlocks,IncompleteBlocks,locationName) %>% 
     spread(locationName,n)
```

Convert the data to “long format” . Remove missing values. “Nest” the data by Trait.

```{r}
traits<-c("DM","MCMDS","logFYLD","logDYLD")
phenos<-phenos %>% 
     # Convert the data to "long format" 
     pivot_longer(cols = all_of(traits), 
                  names_to = "Trait", values_to = "Value") %>%
     # Remove missing values
     filter(!is.na(Value)) %>%
     # Nest the MultiEnvironmentTrial data by trait
     nest(METdata=c(-Trait))
```

```{r}
phenos %>% 
     mutate(N_plots=map_dbl(METdata,nrow))
```

```{r}
phenos$METdata[[1]] %>% head
```

# Comparison of models

```{r}
# grab just one chunk of the data to "experiment" on
METdata<-phenos$METdata[[1]] %>% 
     # sommer doesn't recognize logical T/F variables
     mutate(CompleteBlocks=as.character(CompleteBlocks), 
            IncompleteBlocks=as.character(IncompleteBlocks))
```

How to run a single, simple model, _not_ using future.

This next step takes at least 25 minutes to run on my laptop.
```{r, eval=F}
simple_model<-mmer(Value~yearInLoc,
                   random=~germplasmName + repInTrial + blockInRep,
                   data=METdata, 
                   verbose = FALSE)
```

```{r, eval=F}
library(sommer)
library(future);
plan(multisession)

# simplify: no het. error, no at() variance structure
simple_model %<-% { mmer(Value~yearInLoc,
                         random=~germplasmName + repInTrial + blockInRep,
                         data=METdata, 
                         verbose = FALSE) }

# no at() variance structure, include het. error variance by trial
simple_het_error_model %<-% { mmer(Value~yearInLoc,
                                   random=~germplasmName + repInTrial + blockInRep,
                                   rcov=~vs(ds(studyName),units),
                                   data=METdata, 
                                   verbose = FALSE) }

# add heterogeneous-error variance for each trial (studyName)
conv_het_error_model %<-% { mmer(Value~yearInLoc,
                                 random=~germplasmName + 
                                      vs(at(CompleteBlocks,"TRUE"),repInTrial) + 
                                      vs(at(IncompleteBlocks,"TRUE"),blockInRep),
                                 rcov=~vs(ds(studyName),units),
                                 data=METdata, 
                                 verbose = FALSE) }

# the model I've applied to NextGen predictions the last several years
conv_model %<-% { mmer(Value~yearInLoc,
                       random=~germplasmName + 
                            vs(at(CompleteBlocks,"TRUE"),repInTrial) + 
                            vs(at(IncompleteBlocks,"TRUE"),blockInRep),
                       data=METdata, 
                       verbose = FALSE) }

```

```{r, eval=F}
# Save output of above so the above models don't need to be re-run
## if and when I reformat or otherwise re-knit this doc to HTML
save(conv_model,conv_het_error_model,simple_model,simple_het_error_model,
     file=here::here("output","models_compared.Rdata"))
```

Load the fitted model outputs.

```{r}
# load the output from the models from where I stored it
load(here::here("output","models_compared.Rdata"))
```

Which model fits best? How to compare?

sommer provides `anova()` for pairwise comparisons of models using a likelihood ratio test and by AIC, BIC.

```{r}
anova(conv_model,conv_het_error_model)
```

Here’s a very not-pretty not-high-throughput way to set-up a table comparing the AIC for all 4 models.

```{r}
tibble(Model=c("conv_model","conv_het_error_model",
               "simple_model","simple_het_error_model"),
       AIC=c(conv_model$AIC,conv_het_error_model$AIC,
             simple_model$AIC,simple_het_error_model$AIC)) %>% 
     arrange(AIC)
```

The “simple” heterogeneous error model fits the data best, I’ll go ahead and use it for this demonstration. 

## ACTIVITY

Run the simple model, examine the results. 

Option 1) Use `sommer`. Takes ~9 mins on my laptop.

```{r, eval=F}
simple_model<-mmer(Value~yearInLoc,
                   random=~germplasmName + repInTrial + blockInRep,
                   data=METdata, 
                   verbose = TRUE)
```

Option 2) Use package `lme4`'s `lmer()` function... takes seconds. Same result.

Again, note that `sommer` advantage is flexible modeling. 

Plus, `sommer` can do genomic prediction, `lmer` can't do that.
```{r}
library(lme4)
simple_model_lmer<-lmer(Value~yearInLoc + (1|germplasmName) + (1|repInTrial) + (1|blockInRep),
                        data=METdata)

summary(simple_model_lmer)
```

# Analyze all traits

Next we will fit the “best” model from the previous step, but to all traits.

We have four data chunks set-up in the tibble `phenos`

```{r}
phenos
```

The manual, starting [here](https://wolfemd.github.io/GenomicSelectionManual/preliminary-field-trial-analysis.html#test-code-for-function) walks you through the steps creating and debugging a function to analyze all traits. That function is in an R script called `get_blups.R` in the `code/` folder. 
We will use it below.

```{r}
# Source the R script
source(here::here("code","get_blups.R"))
# Set-up inputs for the function
phenos %<>% 
     mutate(fixedFormula=ifelse(Trait %in% c("logDYLD","logFYLD","logRTNO","logTOPYLD"),
                                "Value ~ yearInLoc + PropNOHAV",
                                "Value ~ yearInLoc"),
            randFormula="~germplasmName + repInTrial + blockInRep",
            rcovFormula="~vs(ds(studyName),units)")
```

This next step took ***45 minutes*** on my laptop.
```{r, eval=F}
starttime<-proc.time()[3]
require(furrr); plan(multisession, workers = 4)
blups<-phenos %>%
     mutate(modelOut=future_pmap(.,get_blups)) %>% 
     unnest(modelOut)
stoptime<-proc.time()[3]; elapsed<-stoptime-starttime; elapsed/60
#  elapsed 
# 44.86077
# Save
saveRDS(blups,file = here::here("output","blups.rds"))
```

```{r}
blups<-readRDS(here::here("output","blups.rds"))
# Print the output, add a few summaries, like H2
blups %>% 
     dplyr::select(Trait,H2,meanHetVarE,meanNreps,outliers) %>% 
     mutate(H2=round(H2,3),
            meanNreps=round(meanNreps,3),
            outliers=map_dbl(outliers,length))
```

Show all the outputs from the model in `blups`
```{r}
blups
```

Show how the BLUPs come out of the function formatted:
```{r}
blups$blups[[1]]
```

That's the end of the core steps.

The rest of the manual chapter includes some exploration of the results _and_ explanation of BLUPs, de-regressed BLUPs and weighted analysis. 

I will try to cover that with some slides to give you intuition.


# Study material

## Learn functions and iteration

At this point, taking a moment to familiarize yourself with some R packages and programming concepts might be adviseable.

1.  Learn to read and write your own functions in R
2.  Learn to iterate procedures in R

### Write functions in R

<https://swcarpentry.github.io/r-novice-inflammation/02-func-R/>

<https://r4ds.had.co.nz/functions.html>

### Iteration in the tidyverse

Packages `purrr` and (for parallel processing) `furrr`.

<https://r4ds.had.co.nz/iteration.html>

<https://www.r-bloggers.com/2021/09/tidy-parallel-processing-in-r-with-furrr/>

<https://purrr.tidyverse.org/>

<https://dcl-prog.stanford.edu/purrr-mutate.html>

<https://dcl-prog.stanford.edu/purrr-parallel.html>

<https://furrr.futureverse.org/>

## Learn mixed-models

<https://si.biostat.washington.edu/sites/default/files/modules/Seattle-SISG-18-IntroQG-Lecture08.pdf>

<https://towardsdatascience.com/how-linear-mixed-model-works-350950a82911>

<https://stats.oarc.ucla.edu/other/mult-pkg/introduction-to-linear-mixed-models/>

[Excellence in Breeding - Basics of GS - 2021 YouTube Video Series with Eduardo Covarrubias](https://www.youtube.com/playlist?list=PLZ0lafzH_UmfHsoePj0t7xQglKSOpzbDp)
